Potential Factors of Decreasing Birth Rates in the US since the Great Recession and Beyond

### 2024-02-12 outputs/paper/paper.qmd
update

According to the Centers for Disease Control and Prevention, the birth rate is defined as "the number of live births per 1,000 population." This rate is calculated by dividing the number of live births in a population in a year by the midyear resident population. Specifically, for census years, rates are calculated based on the unrounded census counts of the resident population as of April 1st. From an economic perspective, birth rates are an important factor that determines the growth of a country. In recent years, the low birth rate has become a conversation topic both nationwide and worldwide. Many countries are addressing this issue of declining birth rates and proposing alternative measures to compensate for the population drop. For instance, Germany's birth rate (births per 1000 people) in 2024 is 9.3, which is lower than Canada's 10.0 and the US's 12.0 (Macrotrends 2024). Germany is taking action by actively recruiting skilled workers from abroad (Grunau 2023) and reforming its citizenship law to allow dual citizenship (Knight 2024) as some measures to combat declining birth rates and overall population. Finally, the birth rate also serves as an indicator of replacement level fertility, which in developed countries can be taken as requiring an average of 2.1 children per woman (Craig 1994). Therefore, birth rates are poised to become a key global topic in the mid to long term.

@@

This paper utilized a replication package of the original study and employed a cleaned data CSV file to reproduce the related figures and tables. This approach enabled us to draw correlations between the dataset and political, economic, and social policy measures that were not covered in the original paper. Furthermore, this paper investigates various socio-cultural aspects, including sexual education, marriage rates, financial stability among women over the age of 30, and the role of technology in healthcare.

-what was found-

@@

## Features
De facto population of women of reproductive age is 15 to 49 years in a country (WHO 2024). The average age of the final menstrual period (menopause) is age 51, women age in 40s, pregnancy is rare but not impossible (asrm 2012). Beginning in 1997, the birth rate for the maternal age group 45-49 includes data for mothers aged 45 and over in the numerator and is based on the population of women aged 45-49 in the denominator (National Center for Health Statistics 2023). Thus the women population in the age group 45-49 is eliminated in this paper although they fall under the reproductive age group category in a country. Birth data is comprised of births per 1,000 women age 15 to 44 over the 40 years period from 1980 to 2020.
For the race and ethnicity, every white refers to non-Hispanic, caucassian population group. Every black refers to non-Hispanic, African American group. Hispanic refers to people with Hispanic ethinicity majority from south America to abberiviate.

@@

## Trend in US Birth Rates
The overall trend fluctuates over time, with the highest number of births recorded in 1990 at a rate of 70.9. The second-highest peaks occurred in 1991 and 2007, with rates of 69.3. These peaks are accompanied by valleys. However, the difference between the birth rates in 1980 and 2020 is 12.6, and the gap between the maximum and minimum points during this period is 15.1. These calculations clearly demonstrate the variation over time and suggest that there may be factors influencing the gradual negative slope of the trend. Figure 2 presents the scatter plot version of Figure 1, providing a clearer representation of the negative trend observed in the line graph. Lastly, the year 2007 has been marked in dotted line to signify the beginning of the fall on both graphs.

@@

## Trends in Birth Rates by Population Subgroup
Trends in Birth Rates by Population Subgroup has two panels specified into two: Group A is designated to five-year age group and Group B is designed to illustrate race and ethnicity (ages 15-44).

@@

Note that the data set is gathered as of 1989 for the race and ethnicity subgroup population, thus the x-axis of the graph starts from 1989. In panel B of @fig-bratepop-graph has three different subgroups. Race and ethnicity is subdivided into three sections, Hispanic, Black, and White. Both white and black are non-Hispanic population. While the white population group shows relatively linear fashion over 30 years, Hispanic population birth groups were 1.7 times and black population birth groups being 1.3 times larger than White population group up until 2007. The birth rates of Hispanic and Black population groups drastically drop over the 10 years from 1989 to 2000. Then another significant drop happened as of 2007 only in Hispanic population group. The change in births per 1,000 women in race and ethnicity population subgroup varies, 7.3, 25.8 and 42.1 in White, Black, and Hispanic group respectively. As of 2012, all race and ethnicity population group illustrate the relatively stable tendency and slowly declining.

@@

Christianity, including protestants and catholic, generally oppose the idea of abortion and view it as morally wrong. They also oppose contraceptive use outside of marriage. Religious group are socially conservative. With this religious reason behind it, the social norm was opposed to the idea of abortion, contraceptive use back in the days which is why this can be a reason why the birth rate is decreasing for all race and ethnicity groups. Shift in religious activities of young Americans. Early 1990s, about 90% of US adults identified as Christians. However, over the time, young Americans are now less likely to become or remain Christian (Pew Research Center 2022). In 2020, unaffiliated group share the greatest part of the American religious landscape in 2020 by parting 23% (PRRI 2020).

@@

## Future Research
Political demographic, immigration law, and enforcement.

>>>>

According to the Centers for Disease Control and Prevention, the birth rate is defined as "the number of live births per 1,000 population." This rate is calculated by dividing the number of live births in a population in a year by the midyear resident population. Specifically, for census years, rates are calculated based on the unrounded census counts of the resident population as of April 1st. From an economic perspective, birth rates are an important factor that determines the growth of a country. In recent years, the low birth rate has become a conversation topic both nationwide and worldwide. Many countries are addressing this issue of declining birth rates and proposing alternative measures to compensate for the population drop. For instance, Germany's birth rate in 2024 is 9.3, which is lower than Canada's 10.0 and the US's 12.0 (Macrotrends 2024). Germany is taking action by actively recruiting skilled workers from abroad (Grunau 2023) and reforming its citizenship law to allow dual citizenship (Knight 2024) as some measures to combat declining birth rates and overall population. Finally, the birth rate also serves as an indicator of replacement level fertility, which in developed countries can be taken as requiring an average of 2.1 children per woman (Craig 1994). Therefore, birth rates are poised to become a key global topic in the mid to long term.

@@

This paper utilized a replication package of the original study and employed a cleaned data CSV file to reproduce the related figures and tables. This approach enabled us to draw correlations between the dataset and

@@

## Features
De facto population of women of reproductive age is 15 to 49 years in a country (WHO 2024). The average age of the final menstrual period (menopause) is age 51, women age in 40s, pregnancy is rare but not impossible (asrm 2012). Beginning in 1997, the birth rate for the maternal age group 45-49 includes data for mothers aged 45 and over in the numerator and is based on the population of women aged 45-49 in the denominator (National Center for Health Statistics 2023). Thus the women population in the age group 45-49 is eliminated in this paper although they fall under the reproductive age group category. Birth data is comprised of births per 1,000 women age 15 to 44 over the 40 years period from 1980 to 2020.
For the race and ethnicity, every white refers to non-Hispanic, caucassian population group. Every black refers to non-Hispanic, African American group. Hispanic refers to people with Hispanic ethinicity majority from south America to abberiviate.

@@

## Trend in US Birth Rates
The overall trend fluctuates over time, with the highest number of births recorded in 1990 at a rate of 70.9. The second-highest peaks occurred in 1991 and 2007, with rates of 69.3. These peaks are accompanied by valleys. However, the difference between the birth rates in 1980 and 2020 is 12.6, and the gap between the maximum and minimum points during this period is 15.1. These calculations clearly demonstrate the variation over time and suggest that there may be factors influencing the gradual negative slope of the trend. @fig-scatter-graph presents the scatter plot version of @fig-bratetrend-graph, providing a clearer representation of the negative trend observed in the line graph. Lastly, the year 2007 has been marked in dotted line to signify the beginning of the fall on both graphs.

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: tbl-polscatter-table
#| tbl-cap: "US presidents"

# Create a dataframe with information about the Presidents and their respective parties
presidents <- data.frame(
  president = c("Ronald Reagan", "George H. W. Bush", "Bill Clinton", "George W. Bush", "Barack Obama", "Donald Trump", "Joe Biden"),
  party = c("Republican", "Republican", "Democratic", "Republican", "Democratic", "Republican", "Democratic"),
  start_year = c(1981, 1989, 1993, 2001, 2009, 2017, 2021),
  end_year = c(1989, 1993, 2001, 2009, 2017, 2021, NA) # NA for current President
)

head(presidents)

```
```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
#| label: fig-polscatter-graph
#| fig-cap: "Trend in US Birth Rates with political party in their presidential mandate"

# Plotting the data with scatter plot and trend lines for each presidential mandate period
ggplot(data, aes(x = year, y = brate_all)) +
  geom_point(color = "pink") +
  geom_smooth(data = subset(data, year >= 1980 & year <= 1981), method = "lm", se = FALSE, color = "blue") +
  geom_smooth(data = subset(data, year >= 1981 & year <= 1989), method = "lm", se = FALSE, color = "red") +
  geom_smooth(data = subset(data, year >= 1989 & year <= 1993), method = "lm", se = FALSE, color = "red") +
  geom_smooth(data = subset(data, year >= 1993 & year <= 2001), method = "lm", se = FALSE, color = "blue") +
  geom_smooth(data = subset(data, year >= 2001 & year <= 2009), method = "lm", se = FALSE, color = "red") +
  geom_smooth(data = subset(data, year >= 2009 & year <= 2017), method = "lm", se = FALSE, color = "blue") +
  geom_smooth(data = subset(data, year >= 2017 & year <= 2021), method = "lm", se = FALSE, color = "red") +
  geom_vline(xintercept = 2007, linetype = "dotted") +
  annotate("text", x = 2009, y = 70, label = "2007", vjust = -0.5, hjust = 0.5, color = "black") +
  geom_hline(yintercept = seq(50, 80, by = 5), color = "lightgrey") +
  scale_y_continuous(limits = c(50, 80), breaks = seq(50, 80, by = 5)) +
  scale_x_continuous(limits = c(1980, 2020), breaks = seq(1980, 2020, by = 1), labels = function(x) ifelse(x %% 5 == 0, x, "")) +
  labs(x = "Year", y = "Births per 1,000 women age 15-44",
       caption = "Source: Birth Rates collected from CDC Vital Statistics Births Reports for 2015, 2019, and 2020.") +
  theme_classic()

## Trends in Birth Rates by Population Subgroup
@fig-bratepop-graph has two panels specified into two: Group A is designated to five-year age group and Group B is designed to illustrate race and ethnicity (ages 15-44).

@@

Note that the data set is gathered as of 1989 for the race and ethnicity subgroup population, thus the x-axis of the graph starts from 1989. In panel B of @fig-bratepop-graph has three different subgroups. Race and ethnicity is subdivided into three sections, Hispanic, Black, and White. Both white and black are non-Hispanic population. While the white population group shows relatively linear fashion over 30 years, Hispanic population birth groups were 1.7 times and black population birth groups being 1.3 times larger than White population group up until 2007. The birth rates of Hispanic and Black population groups drastically drop over the 11 years from 1989 to 2000. Then another significant drop happened as of 2007 only in Hispanic population group. The change in births per 1,000 women in race and ethnicity population subgroup varies, 7.3, 25.8 and 42.1 in White, Black, and Hispanic group respectively. As of 2012, all race and ethnicity population group illustrate the relatively stable tendency and slowly declining.

@@

Another interesting factor could be christianity. Including protestants and catholic, christian generally against with the idea of abortion and view it as morally wrong. They also oppose contraceptive use outside of marriage. Religious group are socially conservative. However, there was a shift in religious activities of young Americans. Early 1990s, about 90% of US adults identified as Christians. However, over the time, young Americans are now less likely to become or remain Christian (Pew Research Center 2022). In 2020, unaffiliated group share the greatest part of the American religious landscape in 2020 by parting 23% (PRRI 2020). This change in religious activity could explain the reason why the birth rate is decreasing for all race and ethnicity groups over the period.

@@

## Future Research
It might be worth investigating some other aspects of the society such as political demographic, immigration law, and enforcement. Based on trend in
